---
title: "Importing external deconvolution results"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Importing external deconvolution results}
  %\VignetteEngine{knitr::knitr}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
```

This vignette shows how to import cell-type estimates obtained with external
deconvolution tools into spacedeconv for spatial visualization and downstream
analysis. As an example, we use omnideconv to run BayesPrism on spatial
transcriptomics data, import the resulting estimates into a SpatialExperiment,
and visualize and cluster spots based on their estimated cell-type composition.

# Install omnideconv

Install **omnideconv in a Conda environment named `r-omnideconv`**, following the
[omnideconv installation instructions](https://omnideconv.org/omnideconv/#installation).
Choose the minimal installation and install the dependencies for **BayesPrism**,
which is the only deconvolution method used here. Keep spacedeconv in its own
`spacedeconv-env` environment.

The code below is intended to be run interactively in two separate R sessions,
using the same working directory. It is not executed when building this vignette.

# Preparing the input files

For demonstration, we load and preprocess the example datasets included with
spacedeconv. For your own analysis, use your spatial transcriptomics data and
an appropriate annotated single-cell reference instead. Run the following in
an R session in `spacedeconv-env`:

```{r save_inputs}
library(spacedeconv)

data("spatial_data_3")
data("single_cell_data_3")

spe <- spacedeconv::preprocess(spatial_data_3)
sce <- spacedeconv::preprocess(single_cell_data_3)

saveRDS(spe, "spatial_data.rds")
saveRDS(SummarizedExperiment::assay(spe, "counts"), "spatial_counts.rds")
saveRDS(sce, "single_cell_reference.rds")
```

Keep these files in the working directory used by both sessions. The supplied
single-cell reference already contains the `celltype_major` annotation column.
Preprocessing retains raw counts, as required for the BayesPrism example.
For your own data, substitute the spatial object, reference, and annotation column.

# Run omnideconv on spatial data

Activate the omnideconv environment in your terminal and start R:

```{bash}
micromamba activate r-omnideconv
R
```

Each spatial spot is supplied as one sample: rows are genes and columns are
spots. For [BayesPrism](https://omnideconv.org/omnideconv/#input-data), use raw
counts for both the spatial data and the annotated single-cell reference.
Gene identifiers and organism must agree between the two datasets.

This example reads `spatial_counts.rds` (a gene-by-spot count matrix) and
`single_cell_reference.rds` (a `SingleCellExperiment` with a `counts` assay and
`celltype_major` annotations).

```{r omnideconv}
library(omnideconv)

spatial_counts <- readRDS("spatial_counts.rds")
reference <- readRDS("single_cell_reference.rds")

fractions <- omnideconv::deconvolute(
  bulk_gene_expression = as.matrix(spatial_counts),
  single_cell_object = reference,
  cell_type_column_name = "celltype_major",
  assay_name = "counts",
  method = "bayesprism",
  n_cores = 1
)

saveRDS(fractions, "bayesprism_results.rds")
```

The result has spots as rows and cell types as columns. Spot IDs are retained
for importing the results. Spatial coordinates are not used in this
omnideconv call; they remain in the original spatial object. Method suitability
for sparse spot data should be assessed for the dataset being analyzed.

# Import into spacedeconv

Switch environments and start a new R session:

```{bash}
micromamba activate spacedeconv-env
R
```

Load the original `SpatialExperiment` and import the results:

```{r import_results}
library(spacedeconv)

spe <- readRDS("spatial_data.rds")
fractions <- readRDS("bayesprism_results.rds")

spe <- import_deconvolution_results(
  spe,
  results = fractions,
  result_name = "bayesprism"
)

available_results(spe, method = "bayesprism")
```

The importer aligns rows by spot ID and stores columns such as
`bayesprism_T_cells` in `colData(spe)`. Results must cover all spots in `spe`;
values must be finite and non-negative. Values are not normalized, and existing
columns are not overwritten. Expression data, coordinates, and images are preserved.

# Visualize and cluster

Use the same result name to plot the imported cell-type estimates or cluster
spots by their estimated composition:

```{r downstream}
plot_spatial(spe, result = "bayesprism", density = FALSE)

spe <- spacedeconv::cluster(
  spe,
  spmethod = "bayesprism",
  method = "hclust",
  dist_method = "euclidean",
  nclusters = 3
)

plot_spatial(spe, result = "cluster_bayesprism_nclusters_3", density = FALSE)
get_cluster_features(spe, clusterid = "cluster_bayesprism_nclusters_3", topn = 3)
```

For datasets with a different sample ID, pass that ID to `plot_spatial()` via
`sample_id`. The same import workflow works with results from other tools:
provide a numeric spot-by-cell-type matrix with row and column names, and choose
a distinct `result_name`.
