---
title: "DecoupleR integration"
output: html_document
date: "2024-02-27"
vignette: >
  %\VignetteIndexEntry{DecoupleR integration}
  %\VignetteEngine{knitr::knitr}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

We provide wrapper functions to spatially quantify Pathway and Transcription Factor activities using DecoupleR.

```{r}
library(spacedeconv)
library(decoupleR)
```

Built-in references can be retrieved via decoupleR by selecting the reference database with the `method` argument: "progeny", "dorothea", or "collectri". Custom reference tables can also be supplied directly, as shown below.

```{r, collapse=TRUE}
ref <- get_decoupleR_reference(method = "progeny", organism = "human")
head(ref)
```

In the next step the activities can be computed.

```{r, eval=TRUE, collapse=TRUE}
data("spatial_data_3")
spatial_data_3 <- preprocess(spatial_data_3)
spatial_data_3 <- spacedeconv::normalize(spatial_data_3)

spe <- compute_activities(spatial_data_3, reference = ref, method = "wmean", assay = "cpm")
```

Custom references can be passed directly to `compute_activities()` as a table
with `source` (gene set or regulator), `target` (gene identifier), and `weight`.
Use weight 1 for unsigned membership, or biologically justified weights when
available. Gene identifiers must match the expression assay row names and
organism. At least five matching genes per set are required, and each
source-target pair should occur only once. Activity scoring methods such as ULM use the supplied weights, whereas
AUCell, FGSEA, GSVA, and ORA ignore them.

```{r, eval=FALSE}
# Illustrative genes only; use a validated signature or database gene set for analysis.
ref <- data.frame(
  source = rep("my_signature", 5),
  target = c("STAT1", "IRF1", "CXCL9", "CXCL10", "IDO1"),
  weight = rep(1, 5)
)
spe <- compute_activities(spatial_data_3, reference = ref, method = "ulm", assay = "cpm")
# Scores are stored in colData(spe), e.g. spe$decoupleR_my_signature.
```

For example, [MSigDB Hallmark gene sets](https://www.gsea-msigdb.org/gsea/msigdb/collections.jsp)
can be retrieved via decoupleR and converted to a custom reference. This example
uses human gene symbols. The initial download retrieves the full MSigDB resource
and may take a while.

```{r, eval=FALSE}
msigdb <- decoupleR::get_resource("MSigDB")
hallmark <- subset(msigdb, collection == "hallmark")
ref <- unique(data.frame(source = hallmark$geneset, target = hallmark$genesymbol))
ref <- ref[!is.na(ref$target) & nzchar(ref$target), ]
ref$weight <- 1

# Optional: keep only selected Hallmark gene sets.
# ref <- ref[ref$source %in% c("HALLMARK_HYPOXIA", "HALLMARK_INTERFERON_GAMMA_RESPONSE"), ]

spe <- compute_activities(spatial_data_3, reference = ref, method = "ulm", assay = "cpm")
# Example result columns:
# spe$decoupleR_HALLMARK_HYPOXIA
# spe$decoupleR_HALLMARK_INTERFERON_GAMMA_RESPONSE
```
