---
title: "Importing external deconvolution results"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Importing external deconvolution results}
  %\VignetteEngine{knitr::knitr}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", error = FALSE)
micromamba <- Sys.which("micromamba")
if (!nzchar(micromamba)) {
  stop("Building this vignette requires micromamba and the r-omnideconv environment with omnideconv and Bisque installed.")
}
# Share temporary input/output files between the two R processes.
build_dir <- tempfile("spacedeconv-external-")
dir.create(build_dir)
knitr::opts_knit$set(root.dir = build_dir)
```

This vignette shows how to import cell-type estimates obtained with external
deconvolution tools into spacedeconv for spatial visualization and downstream
analysis. As an example, we use omnideconv to run Bisque on spatial
transcriptomics data, import the resulting estimates into a SpatialExperiment,
and visualize and cluster spots based on their estimated cell-type composition.

# Install omnideconv

Install **omnideconv in a Conda environment named `r-omnideconv`**, following the
[omnideconv installation instructions](https://omnideconv.org/omnideconv/#installation).
Choose the minimal installation and install the dependencies for **Bisque**,
which is the only deconvolution method used here. Keep spacedeconv in its own
`spacedeconv-env` environment.

The code below can be run interactively in two separate R sessions: one in
`r-omnideconv` and one in `spacedeconv-env`, both using the same working directory.

# Preparing the input files

For demonstration, we load and preprocess the example datasets included with
spacedeconv. For your own analysis, use your spatial transcriptomics data and
an appropriate annotated single-cell reference instead. Run the following in
an R session in `spacedeconv-env`:

```{r save_inputs}
library(spacedeconv)
library(SingleCellExperiment)

data("spatial_data_3")
data("single_cell_data_1")
data("single_cell_data_2")
data("single_cell_data_3")

spe <- spacedeconv::preprocess(spatial_data_3)
# Bisque requires reference cells from at least two donors (three or more recommended).
sce <- cbind(single_cell_data_1, single_cell_data_2, single_cell_data_3)
sce <- spacedeconv::preprocess(sce)

saveRDS(spe, "spatial_data.rds")
saveRDS(SummarizedExperiment::assay(spe, "counts"), "spatial_counts.rds")
saveRDS(sce, "single_cell_reference.rds")
```

Keep these files in the working directory used by both sessions. The supplied
single-cell datasets come from three different donors in the same breast cancer
study. They contain the `celltype_major` annotation and the donor identifier
`orig.ident`. Preprocessing retains raw counts, as required for Bisque.
For your own data, use an appropriate reference with at least two donors
(three or more recommended) and substitute the annotation and donor columns.

# Run omnideconv on spatial data

Activate the omnideconv environment in your terminal and start R:

```{bash, eval=FALSE}
micromamba activate r-omnideconv
R
```

Each spatial spot is supplied as one sample: rows are genes and columns are
spots. For [Bisque](https://omnideconv.org/omnideconv/reference/deconvolute_bisque.html), use raw
counts for both the spatial data and the annotated single-cell reference.
Gene identifiers and organism must agree between the two datasets.

This example reads `spatial_counts.rds` (a gene-by-spot count matrix) and
`single_cell_reference.rds` (a `SingleCellExperiment` with a `counts` assay and
`celltype_major` annotations and `orig.ident` donor IDs).

```{r omnideconv, eval=FALSE}
library(omnideconv)

spatial_counts <- readRDS("spatial_counts.rds")
reference <- readRDS("single_cell_reference.rds")

fractions <- omnideconv::deconvolute(
  bulk_gene_expression = as.matrix(spatial_counts),
  single_cell_object = reference,
  cell_type_column_name = "celltype_major",
  batch_ids = reference$orig.ident,
  assay_name = "counts",
  method = "bisque"
)

saveRDS(fractions, "bisque_results.rds")
```

```{r run_omnideconv, include=FALSE}
# Execute exactly the displayed chunk, without loading omnideconv in this session.
script <- file.path(build_dir, "omnideconv_step.R")
log_file <- file.path(build_dir, "omnideconv.log")
writeLines(knitr::knit_code$get("omnideconv"), script)
status <- withr::with_envvar(
  c(R_HOME = NA, R_LIBS = NA, R_LIBS_USER = NA, R_LIBS_SITE = NA,
    RETICULATE_PYTHON = NA, RETICULATE_PYTHON_ENV = NA),
  system2(micromamba,
    c("run", "-n", "r-omnideconv", "Rscript", shQuote(script)),
    stdout = log_file, stderr = log_file)
)
if (status != 0L || !file.exists("bisque_results.rds")) {
  details <- paste(tail(readLines(log_file, warn = FALSE), 30), collapse = "\n")
  stop("The omnideconv step failed. Check the r-omnideconv environment.\n", details)
}
```

The result has spots as rows and cell types as columns. Spot IDs are retained
for importing the results. Spatial coordinates are not used in this
omnideconv call; they remain in the original spatial object. Method suitability
for sparse spot data should be assessed for the dataset being analyzed.

# Import into spacedeconv

Switch environments and start a new R session:

```{bash, eval=FALSE}
micromamba activate spacedeconv-env
R
```

Load the original `SpatialExperiment` and import the results:

```{r import_results}
library(spacedeconv)

spe <- readRDS("spatial_data.rds")
fractions <- readRDS("bisque_results.rds")

spe <- import_deconvolution_results(
  spe,
  results = fractions,
  result_name = "bisque"
)

available_results(spe, method = "bisque")
```

The importer aligns rows by spot ID and stores columns such as
`bisque_T.cells` in `colData(spe)`. Results must cover all spots in `spe`;
values must be finite and non-negative. Values are not normalized, and existing
columns are not overwritten. Expression data, coordinates, and images are preserved.

# Visualize and cluster

Use the same result name to plot the imported cell-type estimates or cluster
spots by their estimated composition:

```{r downstream, fig.width=10, fig.height=8, out.width="100%", fig.alt="Spatial cell-type fractions and clusters derived from imported Bisque results."}
plot_spatial(spe, result = "bisque", density = FALSE,
             title_size = 10, font_size = 8, legend_size = 12)

spe <- spacedeconv::cluster(
  spe,
  spmethod = "bisque",
  method = "hclust",
  dist_method = "euclidean",
  nclusters = 3
)

plot_spatial(spe, result = "cluster_bisque_nclusters_3", density = FALSE)
get_cluster_features(spe, clusterid = "cluster_bisque_nclusters_3", topn = 3)
```

For datasets with a different sample ID, pass that ID to `plot_spatial()` via
`sample_id`. The same import workflow works with results from other tools:
provide a numeric spot-by-cell-type matrix with row and column names, and choose
a distinct `result_name`.
